Research Notes

### 2019-10-09 writeups/Su_IGW_break.tex
check in exact solved inhomogeneous eq

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2#1}{\partial#2^2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\left\langle#1\right\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \newcommand*{\bm}[1]{\boldsymbol{\mathbf{#1}}}
    \newcommand*{\uv}[1]{\hat{\boldsymbol{\mathbf{#1}}}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}

@@

In the linearized system, the effect of this forcing can be approximated
most clearly as two plane waves propagating away from the forcing zone
\begin{align}
    u_{z}'&(x, z, t) \approx{} \frac{Cgk_{x}^2}{\omega^2}
        \frac{1}{2ik_{z}}\frac{e^{\frac{-(k_{z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}\cos\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}\cos\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ region models an upward propagating IGW wavetrain excited deep
in the atmosphere. A derivation of this result is provided in
\autoref{s:force_solved}.

@@

To solve for the excited wave amplitude in the linear problem, consider the
complexified linearized system of equations, then $\pd{}{t} \to -i\omega,
\pd{}{x} \to ik_x$. Assuming all fields have dependence $u_z(x, z, t) = u_z'(z)
e^{ik_xx - i\omega t}$, this gives:
\begin{align*}
    \rd{u_{z}'}{z} + ik_xu_x' &= 0,\\
    -i\omega u_x' + ik_x \varpi' + gHik_x \Upsilon &= 0,\\
    -i\omega u_{z}' + \rd{\varpi'}{z} + gH\rd{\Upsilon}{z}
        - \frac{\varpi'}{H} &= 0,\\
    -i\omega \Upsilon - \frac{u_{z}'}{H} &=
        Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}.
\end{align*}
We have replaced all partial derivatives with regular derivatives. Compare to
the left hand side of \autoref{se:dedalus_eqs} below. This can be recast solely
in terms of $u_{1z}$ as
\begin{align*}
    \omega^2\p{
        k_x^2u_{z}' + \frac{1}{H}\rd{u_{z}'}{z} - \rtd{u_{z}'}{z}
    } - u_{z}'N^2k_x^2 =
    gk_x^2Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}.
\end{align*}
% In the limit $\sigma \to 0$, the right hand side can be approximated
% $e^{-\frac{(z - z_0)^2}{2\sigma^2}} \approx \sigma\sqrt{2\pi}\delta\p{z - z_0}$.
% Matching this against a singularity in $\ptd{u_z'}{z}$, or a kick in
% $\pd{u_z'}{z}$, gives \autoref{eq:uz_lin}.
We compute the solution to the inhomogeneous ODE by the method of variation of
parameters. The homogeneous solutions are of form $u_{z,\pm}'(z) =
e^{\frac{z}{2H} \pm ik_zz}$ where $k_z$ satisfies the dispersion relation
\autoref{eq:disp_rel}. The Wronskian can be computed
\begin{equation}
    W \equiv \det \begin{vmatrix}
        u_{z,+}' & u_{z,-}' \\[4pt]
        \rd{u_{z,+}'}{z} & \rd{u_{z,-}'}{z}
    \end{vmatrix} = -2ik_ze^{z/H}.
\end{equation}

>>>>

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2#1}{\partial#2^2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\left\langle#1\right\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \newcommand*{\bm}[1]{\boldsymbol{\mathbf{#1}}}
    \newcommand*{\uv}[1]{\hat{\boldsymbol{\mathbf{#1}}}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}
    \DeclareMathOperator*{\erf}{erf}

@@

In the linearized system, the effect of this forcing can be solved exactly. If
we approximate $k_zH \gg 1, \sigma \ll H$, the solution can be approximated as
two plane waves propagating away from the forcing zone
\begin{align}
    u_{z}'&(x, z, t) \approx{} \frac{C}{2k_z}\frac{gk_x^2}{\omega^2}
        e^{-\frac{k_z^2\sigma^2}{2}}
        \sqrt{2\pi \sigma^2} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ region models an upward propagating IGW wavetrain excited deep
in the atmosphere. A derivation of this result is provided in
\autoref{s:force_solved}.

@@

To solve for the excited wave amplitude in the linear problem, consider the
complexified linearized system of equations, then $\pd{}{t} \to -i\omega,
\pd{}{x} \to ik_x$. Assuming all fields have dependence $u_z(x, z, t) = u_z'(z)
e^{ik_xx - i\omega t}$, this gives:
\begin{align*}
    \rd{u_{z}'}{z} + ik_xu_x' &= 0,\\
    -i\omega u_x' + ik_x \varpi' + gHik_x \Upsilon &= 0,\\
    -i\omega u_{z}' + \rd{\varpi'}{z} + gH\rd{\Upsilon}{z}
        - \frac{\varpi'}{H} &= 0,\\
    -i\omega \Upsilon - \frac{u_{z}'}{H} &=
        Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}.
\end{align*}
We have replaced all partial derivatives with regular derivatives. This can be
recast solely in terms of $u_{z}'$ as
\begin{align*}
     \rtd{u_{z}'}{z} - k_x^2u_{z}' - \frac{1}{H}\rd{u_{z}'}{z}
        + u_{z}'\frac{N^2k_x^2}{\omega^2} &=
    -\frac{gk_x^2}{\omega^2}Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}
        .\label{eq:narrow_inhomo}
\end{align*}
The homogeneous solutions are of form $u_{z,\pm}'(z) = e^{\p{\frac{1}{2H} \pm
ik_z}\p{z - z_0}}$ where $k_z$ satisfies the dispersion relation
\autoref{eq:disp_rel}. We compute the solution to the inhomogeneous ODE by the
method of variation of parameters. The Wronskian can be computed
\begin{equation}
    W \equiv \det \begin{vmatrix}
        u_{z,+}' & u_{z,-}' \\[4pt]
        \rd{u_{z,+}'}{z} & \rd{u_{z,-}'}{z}
    \end{vmatrix} = -2ik_ze^{z/H}.
\end{equation}
The general solution is then
\begin{equation}
    u_z' = -u_{z,+}\int \frac{1}{W} u_{z,-}' \p{-\frac{gk_x^2}{\omega^2}
            Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}}\;\mathrm{d}z
        + u_{z,-}\int \frac{1}{W} u_{z,+}' \p{-\frac{gk_x^2}{\omega^2}
            Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}}\;\mathrm{d}z.
\end{equation}
Taking these integrals and applying boundary conditions $u_z'\p{z \to \infty} =
u_{z,+}', u_z'\p{z \to -\infty} = u_{z,-}'$ gives exact solution
\begin{equation}
    u(z) = \frac{C}{2ik_z}\frac{gk_x^2}{\omega^2}
            e^{\frac{\p{\frac{\sigma^2}{2H} \pm ik_z\sigma^2}^2}{2\sigma^2}}
                \sqrt{2\sigma^2} \frac{\sqrt{\pi}}{2}\s{
                    -u_+\p{1 + \erf\p{\xi}}
                    + u_- \p{\erf\p{\xi} - 1}}
\end{equation}
If we are concerned with only $z$ scales significantly larger than $\sigma$,
then we may take $\erf(\xi) \approx \Theta(\xi)$. If we further assume $k_zH \gg
1$ and restore the $e^{ik_xx - i\omega t}$ factor, we recover the cited form
\autoref{eq:uz_lin}
\begin{equation}
    u_{z}'(x, z, t) = -\frac{C}{2ik_z}\frac{gk_x^2}{\omega^2}
        e^{-\frac{k_z^2\sigma^2}{2}}
        \sqrt{2\pi \sigma^2} e^{ik_xx - i\omega t} \times
    \begin{cases}
        e^{\p{\frac{1}{2H} + ik_z}\p{z - z_0} + \frac{k_z\sigma^2}{2H}}
            & z > z_0,\\
        e^{\p{\frac{1}{2H} - ik_z}\p{z - z_0} - \frac{k_z\sigma^2}{2H}}
            & z > z_0.
    \end{cases}
\end{equation}
We use the plane wave approximation in our fitting routines since we generally
fit many $\sigma$ away from $z_0$, and orthogonality with other modes is
essential for a reliable fit.

### 2019-10-09 writeups/notes.tex
check in exact solved inhomogeneous eq

    \let\Re\undefined
    \let\Im\undefined
    \DeclareMathOperator{\Res}{Res}
    \DeclareMathOperator{\Re}{Re}
    \DeclareMathOperator{\Im}{Im}
    \DeclareMathOperator{\Log}{Log}
    \DeclareMathOperator{\Arg}{Arg}
    \DeclareMathOperator{\Tr}{Tr}
    \DeclareMathOperator{\E}{E}
    \DeclareMathOperator{\Var}{Var}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator{\sgn}{sgn}
    \DeclareMathOperator{\diag}{diag\;}

@@

>>>>

    \let\Re\undefined
    \let\Im\undefined
    \DeclareMathOperator{\Res}{Res}
    \DeclareMathOperator{\erf}{erf}
    \DeclareMathOperator{\Re}{Re}
    \DeclareMathOperator{\Im}{Im}
    \DeclareMathOperator{\Log}{Log}
    \DeclareMathOperator{\Arg}{Arg}
    \DeclareMathOperator{\Tr}{Tr}
    \DeclareMathOperator{\E}{E}
    \DeclareMathOperator{\Var}{Var}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator{\sgn}{sgn}
    \DeclareMathOperator{\diag}{diag\;}

@@

\subsection{Narrow-Band Forcing: Exact}

We use variation of parameters. We start with PDE in $\varpi, \Upsilon$
coordinates and transform to ODE
\begin{align}
    \rd{u_{z}'}{z} + ik_xu_x' &= 0,\nonumber\\
    -i\omega u_x' + ik_x \varpi' + gHik_x \Upsilon &= 0,\nonumber\\
    -i\omega u_{z}' + \rd{\varpi'}{z} + gH\rd{\Upsilon}{z}
        - \frac{\varpi'}{H} &= 0,\nonumber\\
    -i\omega \Upsilon - \frac{u_{z}'}{H} &=
        Ce^{-\frac{(z - z_0)^2}{2\sigma^2}},\nonumber\\
     \rtd{u_{z}'}{z} - k_x^2u_{z}' - \frac{1}{H}\rd{u_{z}'}{z}
        + u_{z}'\frac{N^2k_x^2}{\omega^2} &=
    -\frac{gk_x^2}{\omega^2}Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}
        .\label{eq:narrow_inhomo}
\end{align}
We have divided through by $e^{i(k_xx - \omega t)}$ and taken $u_z'(x, z, t)
\Rightarrow u_z'(z)$. The inhomogeneous solutions are most easily written
\begin{equation}
    u_{z,\pm}'(z) = e^{\p*{\frac{1}{2H} \pm ik_z}\p*{z - z_0}},
\end{equation}
in which case the Wronskian is
\begin{equation}
    W \equiv \det \begin{vmatrix}
        u_{z,+}' & u_{z,-}' \\[4pt]
        \rd{u_{z,+}'}{z} & \rd{u_{z,-}'}{z}
    \end{vmatrix} = -2ik_ze^{z/H}.
\end{equation}

The general solution to \autoref{eq:narrow_inhomo} is then given $u(z) = A_+u_+
+ A_-u_-$ where
\begin{align}
    A_{\pm} &= \mp \int \frac{1}{W} u_{\mp} \p*{-\frac{gk_x^2}{\omega^2}
            Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}}\;\mathrm{d}z,\nonumber\\
        &= \mp \frac{1}{2ik_z}\frac{gk_x^2}{\omega^2}\int \exp\s*{
            \p*{-\frac{1}{2H} \mp ik_z}z' - \frac{z'^2}{2\sigma^2}}
            \;\mathrm{d}z',\nonumber\\
        &= \mp \frac{1}{2ik_z}\frac{gk_x^2}{\omega^2}\int \exp\s*{
            -\frac{z'^2 + \p*{\frac{\sigma^2}{H} \pm 2ik_z\sigma^2}z'
            }{2\sigma^2}}\;\mathrm{d}z',\nonumber\\
        &= \mp \frac{1}{2ik_z}\frac{gk_x^2}{\omega^2}\int \exp\s*{
            -\frac{\p*{z' + \frac{\sigma^2}{2H} \pm ik_z\sigma^2}^2
                - \p*{\frac{\sigma^2}{2H} \pm ik_z\sigma^2}^2
            }{2\sigma^2}}\;\mathrm{d}z',\nonumber\\
        &= \mp \frac{1}{2ik_z}\frac{gk_x^2}{\omega^2}
            e^{\frac{\p*{\frac{\sigma^2}{2H} \pm ik_z\sigma^2}^2}{2\sigma^2}}
                \sqrt{2\sigma^2} \int e^{-\xi^2}\;\mathrm{d}\xi,\nonumber\\
        &= \mp \frac{1}{2ik_z}\frac{gk_x^2}{\omega^2}
            e^{\frac{\p*{\frac{\sigma^2}{2H} \pm ik_z\sigma^2}^2}{2\sigma^2}}
                \sqrt{2\sigma^2} \frac{\sqrt{\pi}}{2}\p*{1 + C +
                    \erf\p*{\frac{z - z_0}{\sqrt{2\sigma^2}}
                        + \frac{\sigma}{2H\sqrt{2}}
                        \pm \frac{ik_z\sigma}{\sqrt{2}}}}
\end{align}
We have made $z' \equiv z - z_0, \xi = \s*{ \frac{z - z_0}{\sqrt{2\sigma^2}} +
\frac{\sigma}{2H\sqrt{2}} \pm \frac{ik_z\sigma}{\sqrt{2}}}$, and $\erf(z) =
\frac{2}{\sqrt{\pi}}\int\limits_0^z e^{-t^2}\;\mathrm{d}t$. $C$ is a constant of
integration. Since we want $u(z\to \infty) = u_+, u(z \to -\infty) = u_-$, we
can quickly see that $C_+ = 0, C_- = -2$. Thus, the specific solution can be
written down
\begin{equation}
    u(z) = \frac{1}{2ik_z}\frac{gk_x^2}{\omega^2}
            e^{\frac{\p*{\frac{\sigma^2}{2H} \pm ik_z\sigma^2}^2}{2\sigma^2}}
                \sqrt{2\sigma^2} \frac{\sqrt{\pi}}{2}\s*{
                    -u_+\p*{1 + \erf\p*{\xi}}
                    + u_- \p*{\erf\p*{\xi} - 1}}
\end{equation}
In the limit where $H \to \infty$ and $\erf(\xi) \approx \Theta(\xi)$, we
recover
\begin{equation}
    u(z) = -\frac{1}{2ik_z}\frac{gk_x^2}{\omega^2}e^{-\frac{k_z^2\sigma^2}{2}}
        \sqrt{2\pi \sigma^2} \times
    \begin{cases}
        e^{\p*{\frac{1}{2H} + ik_z}\p*{z - z_0} + \frac{k_z\sigma^2}{2H}}
            & z > z_0,\\
        e^{\p*{\frac{1}{2H} - ik_z}\p*{z - z_0} - \frac{k_z\sigma^2}{2H}}
            & z > z_0.
    \end{cases}
\end{equation}
